Cost- and Deadline-Constrained Scheduling of Scientific Workflow Ensembles in IaaS Clouds

### 2012-04-23 paper/paper.tex
better Fig 5
Merge SPDS with DPDS, change algorithms' font sizes

This section describes several algorithms were developed to schedule and provision resources for ensembles of workflows on the cloud under budget and deadline constraints.

\subsection{Static Provisioning Dynamic Scheduling (SPDS) }
\label{sec:spds}

The simplest strategy for executing ensembles of workflows in the cloud is to provision resources statically and schedule workflow tasks dynamically. Given a budget in dollars $b$, deadline in hours $d$, and the hourly price of a VM in dollars $p$, it is easy to calculate the number of VMs, $N_{VM}$, to provision so that the entire budget is consumed before the deadline:
%
\begin{equation}
\label{eq:static-plan}
N_{VM} = \lceil b / (d * p) \rceil
\end{equation}
%
The SPDS algorithm statically provisions $N_{VM}$ VMs at the start of the ensemble execution and keeps them running until the deadline is reached, or the budget runs out. This provisioning plan has the advantage that it minimizes the number of provisioning and de-provisioning requests.

Once the VMs are provisioned, the tasks are mapped onto idle VMs using the dynamic priority-based scheduling procedure shown in Algorithm~\ref{alg:ds}.

\begin{algorithm}[tb]
\caption{Priority-based scheduling algorithm for SPDS}
\label{alg:ds}
\begin{algorithmic}[1]
\Procedure{Schedule}{}
    \State $P\gets$ empty priority queue
	\State $IdleVMs\gets$ set of idle VMs
	\For{root task $t$ in all workflows}
    	\State \Call{Insert}{$t,P$} 
    \EndFor
    \While{deadline not reached}
    	\While{$IdleVMs \neq \emptyset\ and\ P \neq \emptyset$}
    		\State $v\gets$ \Call{SelectRandom}{$IdleVMs$}
    		\State $t\gets$ \Call{Pop}{$P$}
    		\State \Call{Submit}{$t,v$}
    	\EndWhile
    	\State Wait for task $t$ to finish on VM $v$
    	\State Update $P$ with ready children of $t$
		\State \Call{Insert}{$v,IdleVMs$}
    \EndWhile
\EndProcedure
\end{algorithmic} 
\end{algorithm}

@@

This algorithm guarantees that tasks from lower priority workflows are always deferred when higher-priority tasks are available, but lower-priority tasks can still occupy idle VMs when higher-priority tasks are not available. However, because there is no preemption, long-running low-priority tasks may delay the execution of higher-priority tasks. In addition, tasks from low priority workflows may be executed even though there is no chance that those workflows will be completed within the current budget and deadline. Figure~\ref{fig:spds-example} shows an example schedule generated using the SPDS algorithm. The figure illustrates how tasks from lower priority workflows backfill idle VMs when tasks from higher priority workflows are not available.

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/spds-gantt}
  \caption[Example schedule generated by the SPDS algorithm.]{Example schedule generated by the SPDS algorithm. Each row is a different VM. Tasks are represented as boxes and are colored by workflow.}
  \label{fig:spds-example}
\end{figure}

\subsection{Dynamic Provisioning Dynamic Scheduling (DPDS)}

The static provisioning approach of SPDS may perform poorly when resource demand changes during ensemble execution. In those cases, SPDS may either over-provision, wasting the budget on idle VMs, or under-provision, using up the deadline while starving the application of usable VMs. One of the advantages of elasticity in cloud computing is that the number of resources can be changed to meet the demands of an application. In policy-based provisioning (or auto-scaling) systems, metrics such as resource utilization or task queue length are monitored to estimate application demand. This is the approach used by the Dynamic Provisioning Dynamic Scheduling (DPDS) algorithm to determine when to provision and de-provision VMs. The algorithm periodically computes resource utilization using the percentage of idle VMs over time, and adjusts the number of VMs if the utilization is above or below given thresholds. Because it is assumed that VMs are billed by the hour, DPDS only considers VMs that are approaching their hourly billing cycle when deciding which VMs to terminate. The DPDS algorithm is shown in Algorithm~\ref{alg:prov}.

\begin{algorithm}[tb]
\caption{Dynamic provisioning algorithm for DPDS}
\label{alg:prov}
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization threshold; $u_l$: lower utilization
threshold; $v_{max}$: maximum number of VMs
\Procedure{Provision}{}
  \State $V_R\gets$ set of running VMs
    \State $V_C\gets$ set of VMs completing billing cycle
    \State $V_T\gets \emptyset$ \Comment{set of VMs to terminate}
    \State $n_T\gets 0$ \Comment{number of VMs to terminate}
    \If{$ b - c < |V_C| * p\ or\ t > d $ }
      \State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$
      \State $V_T\gets$ select $n_T$ VMs to terminate from $V_C$
      \State \Call{Terminate}{$V_T$} \label{l:terminate1}
    \Else 
    \State $u\gets$ current VM utilization
      \If{$u>u_h$ and $|V_R| < v_{max}*N_{VM}$}
        \State \Call{Start}{$new\ VM$}
      \ElsIf{$u<u_l$}
        \State $V_I\gets$ set of idle VMs
        \State $n_T\gets \lceil|V_I|/2\rceil$ \label{l:nT2}
      \State $V_T\gets$ select $n_T$ VMs to terminate from $V_I$
        \State \Call{Terminate}{$V_T$} \label{l:terminate2}
      \EndIf 
    \EndIf
\EndProcedure
\end{algorithmic} 
\end{algorithm}

@@

\begin{algorithm}[tb]
\caption{Workflow admission algorithm for WA-DPDS}
\label{alg:wa-dpds}
\begin{algorithmic}[1]
\Require $w$: workflow; $b$: budget; $c$: current cost
\Procedure{Admit}{$w,b,c$}
    \State $r_n\gets b-c$ \Comment{Budget remaining for new VMs}
    \State $r_c\gets $cost committed to VMs that are running
    \State $r_a\gets $cost to complete workflows previously admitted
  \State $r_m\gets 0.1$ \Comment{Safety margin}
  \State $r_b\gets r_n+r_c-r_a-r_m$ \Comment{Budget remaining}
  \State $c_w\gets$ \Call{EstimateCost}{w}
    \If{$c_w<r_b$}
      \State \textbf{return} $TRUE$
    \Else
      \State \textbf{return} $FALSE$
  \EndIf       
\EndProcedure
\end{algorithmic} 
\end{algorithm}

@@

\begin{algorithm}[tb]
\caption{Ensemble planning algorithm for SPSS}
\label{alg:admit}
\begin{algorithmic}[1]
\Require $W$: workflow ensemble; $b$: budget; $d$: deadline
\Ensure Schedule as much of $W$ as possible given $b$ and $d$
\Procedure{PlanEnsemble}{$W,b,d$}
    \State $P\gets \emptyset$ \Comment{Current plan}
    \State $A\gets \emptyset$ \Comment{Set of admitted DAGs}
    \For{$w\ in\ W$}
        \State $P^\prime \gets$\ \Call{PlanWorkflow}{$w,P,d$}
        \If{$Cost(P^\prime) \le b$}
            \State $P\gets\ P^\prime$ \Comment{Accept new plan}
            \State $A \gets A\ +\ w$ \Comment{Admit w}
        \EndIf
    \EndFor
    \State \textbf{return} $P,A$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[tb]
\caption{Workflow planning algorithm for SPSS}
\label{alg:planworkflow}
\begin{algorithmic}[1]
\Require $w$: workflow; $P$: current plan; $d$: deadline
\Ensure Create plan for $w$ that minimizes cost and meets deadline $d$
\Procedure{PlanWorkflow}{$w,P,d$}
    \State $P^\prime\gets$ copy of $P$
    \State \Call{DeadlineDistribution}{w,d}
    \For{$t\ in\ w\ sorted\ by\ DL(t)$}
        \State $v \gets$ VM that minimizes cost and start time of t
        \If{$FinishTime(t,v) < DL(t)$}
            \State Schedule(t,v)
        \Else
            \State Provision a new VM v
            \State Schedule(t,v)
        \EndIf
    \EndFor
    \State \textbf{return} $P^\prime$
\EndProcedure
\end{algorithmic}
\end{algorithm}

>>>>

This section describes three algorithms were developed to schedule and provision resources for ensembles of workflows on the cloud under budget and deadline constraints.

\subsection{Dynamic Provisioning Dynamic Scheduling (DPDS)}
\label{sec:dpds}

DPDS is an online algorithm that provisions resources and schedules tasks at runtime. It consists of two main parts: a provisioning procedure, and a scheduling procedure.

DPDS' provisioning procedure is based on resource utilization. DPDS starts with a fixed number of resources calculated based on the available time and budget, and adjusts the number of resources according to how well they are utilized by the application.

Given a budget in dollars $b$, deadline in hours $d$, and the hourly price of a VM in dollars $p$, it is easy to calculate the number of VMs, $N_{VM}$, to provision so that the entire budget is consumed before the deadline:
%
\begin{equation}
\label{eq:static-plan}
N_{VM} = \lceil b / (d * p) \rceil
\end{equation}
%
The DPDS algorithm provisions $N_{VM}$ VMs at the start of the ensemble execution.

As the application runs, the DPDS algorithm periodically computes resource utilization using the percentage of idle VMs over time, and adjusts the number of VMs if the utilization is above or below given thresholds. Because it is assumed that VMs are billed by the hour, DPDS only considers VMs that are approaching their hourly billing cycle when deciding which VMs to terminate. This dynamic provisioning algorithm is shown in Algorithm~\ref{alg:prov}.

\begin{algorithm}[tb]
\caption{Dynamic provisioning algorithm for DPDS}
\label{alg:prov}
{\small
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization threshold; $u_l$: lower utilization
threshold; $v_{max}$: maximum number of VMs
\Procedure{Provision}{}
  \State $V_R\gets$ set of running VMs
    \State $V_C\gets$ set of VMs completing billing cycle
    \State $V_T\gets \emptyset$ \Comment{set of VMs to terminate}
    \State $n_T\gets 0$ \Comment{number of VMs to terminate}
    \If{$ b - c < |V_C| * p\ or\ t > d $ }
      \State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$
      \State $V_T\gets$ select $n_T$ VMs to terminate from $V_C$
      \State \Call{Terminate}{$V_T$} \label{l:terminate1}
    \Else 
    \State $u\gets$ current VM utilization
      \If{$u>u_h$ and $|V_R| < v_{max}*N_{VM}$}
        \State \Call{Start}{$new\ VM$}
      \ElsIf{$u<u_l$}
        \State $V_I\gets$ set of idle VMs
        \State $n_T\gets \lceil|V_I|/2\rceil$ \label{l:nT2}
      \State $V_T\gets$ select $n_T$ VMs to terminate from $V_I$
        \State \Call{Terminate}{$V_T$} \label{l:terminate2}
      \EndIf 
    \EndIf
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

In order to schedule individual workflow tasks onto available VMs, DPDS uses the dynamic, priority-based scheduling procedure shown in Algorithm~\ref{alg:ds}.

\begin{algorithm}[tb]
\caption{Priority-based scheduling algorithm for DPDS}
\label{alg:ds}
{\small
\begin{algorithmic}[1]
\Procedure{Schedule}{}
    \State $P\gets$ empty priority queue
	\State $IdleVMs\gets$ set of idle VMs
	\For{root task $t$ in all workflows}
    	\State \Call{Insert}{$t,P$} 
    \EndFor
    \While{deadline not reached}
    	\While{$IdleVMs \neq \emptyset\ and\ P \neq \emptyset$}
    		\State $v\gets$ \Call{SelectRandom}{$IdleVMs$}
    		\State $t\gets$ \Call{Pop}{$P$}
    		\State \Call{Submit}{$t,v$}
    	\EndWhile
    	\State Wait for task $t$ to finish on VM $v$
    	\State Update $P$ with ready children of $t$
		\State \Call{Insert}{$v,IdleVMs$}
    \EndWhile
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

@@

The DPDS algorithm guarantees that tasks from lower priority workflows are always deferred when higher-priority tasks are available, but lower-priority tasks can still occupy idle VMs when higher-priority tasks are not available. However, because there is no preemption, long-running low-priority tasks may delay the execution of higher-priority tasks. In addition, tasks from low priority workflows may be executed even though there is no chance that those workflows will be completed within the current budget and deadline. Figure~\ref{fig:spds-example} shows an example schedule generated using the SPDS algorithm. The figure illustrates how tasks from lower priority workflows backfill idle VMs when tasks from higher priority workflows are not available.

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/spds-gantt}
  \caption[Example schedule generated by the DPDS algorithm.]{Example schedule generated by the DPDS algorithm. Each row is a different VM. Tasks are represented as boxes and are colored by workflow.}
  \label{fig:spds-example}
\end{figure}

@@

\begin{algorithm}[tb]
\caption{Workflow admission algorithm for WA-DPDS}
\label{alg:wa-dpds}
{\small
\begin{algorithmic}[1]
\Require $w$: workflow; $b$: budget; $c$: current cost
\Procedure{Admit}{$w,b,c$}
    \State $r_n\gets b-c$ \Comment{Budget remaining for new VMs}
    \State $r_c\gets $cost committed to VMs that are running
    \State $r_a\gets $cost to complete workflows previously admitted
  \State $r_m\gets 0.1$ \Comment{Safety margin}
  \State $r_b\gets r_n+r_c-r_a-r_m$ \Comment{Budget remaining}
  \State $c_w\gets$ \Call{EstimateCost}{w}
    \If{$c_w<r_b$}
      \State \textbf{return} $TRUE$
    \Else
      \State \textbf{return} $FALSE$
  \EndIf       
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

@@

\begin{algorithm}[tb]
\caption{Ensemble planning algorithm for SPSS}
\label{alg:admit}
{\small
\begin{algorithmic}[1]
\Require $W$: workflow ensemble; $b$: budget; $d$: deadline
\Ensure Schedule as much of $W$ as possible given $b$ and $d$
\Procedure{PlanEnsemble}{$W,b,d$}
    \State $P\gets \emptyset$ \Comment{Current plan}
    \State $A\gets \emptyset$ \Comment{Set of admitted DAGs}
    \For{$w\ in\ W$}
        \State $P^\prime \gets$\ \Call{PlanWorkflow}{$w,P,d$}
        \If{$Cost(P^\prime) \le b$}
            \State $P\gets\ P^\prime$ \Comment{Accept new plan}
            \State $A \gets A\ +\ w$ \Comment{Admit w}
        \EndIf
    \EndFor
    \State \textbf{return} $P,A$
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

@@

\begin{algorithm}[tb]
\caption{Workflow planning algorithm for SPSS}
\label{alg:planworkflow}
{\small
\begin{algorithmic}[1]
\Require $w$: workflow; $P$: current plan; $d$: deadline
\Ensure Create plan for $w$ that minimizes cost and meets deadline $d$
\Procedure{PlanWorkflow}{$w,P,d$}
    \State $P^\prime\gets$ copy of $P$
    \State \Call{DeadlineDistribution}{w,d}
    \For{$t\ in\ w\ sorted\ by\ DL(t)$}
        \State $v \gets$ VM that minimizes cost and start time of t
        \If{$FinishTime(t,v) < DL(t)$}
            \State Schedule(t,v)
        \Else
            \State Provision a new VM v
            \State Schedule(t,v)
        \EndIf
    \EndFor
    \State \textbf{return} $P^\prime$
\EndProcedure
\end{algorithmic}
}
\end{algorithm}
